The Unicode Cookbook \linebreak for Linguists

### 2018-06-23 book/chapters/implementation.tex
ready for release, except doi

You can also install the \texttt{segments} package from the GitHub repository:\footnote{\url{https://github.com/cldf/segments}}

@@

\subsection*{Application programming interface}
The \texttt{segments} API can be accessed by importing the package into Python or by writing a Python script. Here is an example of how to import the libraries, create a tokenizer object, tokenize a string, and create an orthography profile.

@@

\noindent The \texttt{characters} function will segment a string at Unicode code points.

% \lstset{extendedchars=false, escapeinside=**}
\begin{lstlisting}[basicstyle=\myfont, extendedchars=false, escapeinside={(*@}{@*)}]
>>> t = Tokenizer()
>>> result = t.characters('(*@ĉháɾã̌ctʼɛ↗ʐː| k͡p@*)')
>>> print(result)
>>> '(*@c ̂ h a ́ ɾ a ̃ ̌ c t ʼ ɛ ↗ ʐ ː | # k ͡ p@*)'
\end{lstlisting}

\noindent The \texttt{grapheme\_clusters} function will segment text at the Unicode Extended Grapheme Cluster boundaries.\footnote{\url{http://www.unicode.org/reports/tr18/tr18-19.html\#Default_Grapheme_Clusters}}

\begin{lstlisting}[extendedchars=false, escapeinside={(*@}{@*)}, basicstyle=\myfont]
>>> result = t.grapheme_clusters('(*@ĉháɾã̌ctʼɛ↗ʐː| k͡p@*)')
>>> print(result)
>>> '(*@ĉ h á ɾ ã̌ c t ʼ ɛ ↗ ʐ ː | # k͡ p@*)'
\end{lstlisting}

\noindent The \texttt{grapheme\_clusters} function is the default segmentation algorithm for the \texttt{segments.Tokenizer}. It is useful when you encounter a text that you want to tokenize to identify orthographic or transcription elements.

\begin{lstlisting}[extendedchars=false, escapeinside={(*@}{@*)}, basicstyle=\myfont]
>>> result = t('(*@ĉháɾã̌ctʼɛ↗ʐː| k͡p@*)')
>>> print(result)
>>> '(*@ĉ h á ɾ ã̌ c t ʼ ɛ ↗ ʐ ː | # k͡ p@*)'
\end{lstlisting}

\noindent The \texttt{ipa} parameter forces grapheme segmentation for IPA strings.

\begin{lstlisting}[extendedchars=false, escapeinside={(*@}{@*)}, basicstyle=\myfont]
>>> result = t('(*@ĉháɾã̌ctʼɛ↗ʐː| k͡p@*)', ipa=True)
>>> print(result)
>>> '(*@ĉ h á ɾ ã̌ c tʼ ɛ ↗ ʐː | # k͡p@*)'
\end{lstlisting}

\noindent We can also load an orthography profile and tokenize an input string with it. In the data directory, we've placed an example orthography profile. Let's have a look at it using \texttt{more} on the command line.

@@

\noindent An orthography profile is a delimited UTF-8 text file (here we use tab as a delimiter for reading ease). The first column must be labelled \texttt{Grapheme}. Each row in the \texttt{Grapheme} column specifies graphemes that may be found in the orthography of the input text. In this example, we provide additional columns IPA and XSAMPA, which are mappings from our graphemes to their IPA and XSAMPA transliterations. The final column \texttt{COMMENT} is for comments; if you want to use a tab ``quote that     string''!

@@

\noindent This example shows how we can tokenize input text into our orthographic specification. We can also segment graphemes and transliterate them into other formats, which is useful when you have sources with different orthographies, but you want to be able to compare them using a single representation like IPA.

\begin{lstlisting}[extendedchars=false, escapeinside={(*@}{@*)}, basicstyle=\myfont]
>>> t.transform('(*@aabchonn-ih@*)', 'IPA')
>>> '(*@aː b tʃ õ n í@*)'
\end{lstlisting}

@@

\noindent And we can transliterate to XSAMPA.

\begin{lstlisting}[language=bash, basicstyle=\myfont, extendedchars=false, escapeinside={(*@}{@*)}]
  $ cat sources/german.txt | segments (*@--@*)mapping=XSAMPA 
	(*@--@*)profile=data/german(*@-orthography(*@-@*)profile.tsv tokenize

>>>>

You can also install the \texttt{segments} package from the GitHub repository,\footnote{\url{https://github.com/cldf/segments}} in particular if you would like to contribute to the code base:\footnote{\url{https://github.com/cldf/segments/blob/master/CONTRIBUTING.md}}

@@

\subsection*{Application programming interface}
The \texttt{segments} API can be accessed by importing the package into Python. Here is an example of how to import the library, create a tokenizer object, tokenize a string, and create an orthography profile. Begin by importing the \texttt{Tokenizer} from the \texttt{segments} library.

@@

\noindent Next, instantiate a tokenizer object, which takes optional arguments for an orthography profile and an orthography profile rules file.

\begin{lstlisting}[basicstyle=\myfont]
>>> t = Tokenizer()
\end{lstlisting}

\noindent The default tokenization strategy is to segment some input text at the Unicode Extended Grapheme Cluster boundaries,\footnote{\url{http://www.unicode.org/reports/tr18/tr18-19.html\#Default_Grapheme_Clusters}} and to return, by default, a space-delimited string of graphemes. White space between input string sequences is by default separated by a hash symbol <\#>, which is a linguistic convention used to denote word boundaries. The default grapheme tokenization is useful when you encounter a text that you want to tokenize to identify potential orthographic or transcription elements.

\begin{lstlisting}[extendedchars=false, escapeinside={(*@}{@*)}, basicstyle=\myfont]
>>> result = t('(*@ĉháɾã̌ctʼɛ↗ʐː| k͡p@*)')
>>> print(result)
>>> '(*@ĉ h á ɾ ã̌ c t ʼ ɛ ↗ ʐ ː | \# k͡ p@*)'
\end{lstlisting}

\begin{lstlisting}[extendedchars=false, escapeinside={(*@}{@*)}, basicstyle=\myfont]
>>> result = t('(*@ĉháɾã̌ctʼɛ↗ʐː| k͡p@*)', segment_separator='(*@-@*)')
>>> print(result)
>>> '(*@ĉ-h-á-ɾ-ã̌-c-t-ʼ-ɛ-↗-ʐ-ː-| \# k͡ -p@*)'
\end{lstlisting}

\begin{lstlisting}[extendedchars=false, escapeinside={(*@}{@*)}, basicstyle=\myfont, showstringspaces=false]
>>> result = t('(*@ĉháɾã̌ctʼɛ↗ʐː| k͡p@*)', separator=' // '))
>>> print(result)
>>> '(*@ĉ h á ɾ ã̌ c t ʼ ɛ ↗ ʐ ː | // k͡ p@*)'
\end{lstlisting}

\noindent The optional \texttt{ipa} parameter forces grapheme segmentation for IPA strings.\footnote{\url{https://en.wikipedia.org/wiki/International\_Phonetic\_Alphabet}} Note here that Unicode Spacing Modifier Letters,\footnote{\url{https://en.wikipedia.org/wiki/Spacing\_Modifier\_Letters}} such as <ː> and <\dia{0361}{\large\fontspec{CharisSIL}◌}>, will be segmented together with base characters (although you might need orthography profiles and rules to correct these in your input source; see Section \ref{pitfall-different-notions-of-diacritics} for details).

\begin{lstlisting}[extendedchars=false, escapeinside={(*@}{@*)}, basicstyle=\myfont]
>>> result = t('(*@ĉháɾã̌ctʼɛ↗ʐː| k͡p@*)', ipa=True)
>>> print(result)
>>> '(*@ĉ h á ɾ ã̌ c t ʼ ɛ ↗ ʐː | \# k͡p@*)'
\end{lstlisting}

\noindent You can also load an orthography profile and tokenize input strings with it. In the data directory,\footnote{https://github.com/unicode-cookbook/recipes/tree/master/Basics/data} we've placed an example orthography profile. Let's have a look at it using \texttt{more} on the command line.

@@

\noindent An orthography profile is a delimited UTF-8 text file (here we use tab as a delimiter for reading ease). The first column must be labeled \texttt{Grapheme}, as discussed in Section \ref{formal-specification-of-orthography-profiles}. Each row in the \texttt{Grapheme} column specifies graphemes that may be found in the orthography of the input text. In this example, we provide additional columns \texttt{IPA} and \texttt{XSAMPA}, which are mappings from our graphemes to their IPA and X-SAMPA transliterations. The final column \texttt{COMMENT} is for comments; if you want to use a tab ``quote that     string''!

@@

\noindent This example shows how we can tokenize input text into our orthographic specification. We can also segment graphemes and transliterate them into other forms, which is useful when you have sources with different orthographies, but you want to be able to compare them using a single representation like IPA or X-SAMPA.

\begin{lstlisting}[extendedchars=false, escapeinside={(*@}{@*)}, basicstyle=\myfont]
>>> t('(*@aabchonn-ih@*)', column='IPA')
>>> '(*@aː b tʃ õ n í@*)'
\end{lstlisting}

@@

\noindent And we can transliterate to X-SAMPA.

\begin{lstlisting}[language=bash, basicstyle=\myfont, extendedchars=false, escapeinside={(*@}{@*)}]
  $ cat sources/german.txt | segments (*@--@*)mapping=XSAMPA 
	(*@--@*)profile=data/german(*@-@*)orthography(*@-@*)profile.tsv tokenize

### 2018-06-23 book/chapters/pitfalls.tex
various fixes
ready for release, except doi

We also suggest installing at least one \textsc{fall-back
font}, which provides glyphs that show the user some information about
the underlying encoded character. Apple computers have such a font (which is
invisible to the user), which is designed by Michael Everson and made available
for other systems through the Unicode Consortium.\footnote{
\url{http://www.unicode.org/policies/lastresortfont\_eula.html}} Further, the
\textsc{GNU Unifont} is a clever way to produce bitmaps approximating the
intended glyph of each available character.\footnote{\url{http://unifoundry.com/unifont.html}} Finally,
SIL International provides a \textsc{SIL Unicode BMP Fallback
Font}. This font does not show a real 
glyph, but instead shows the hexadecimal code inside a box
for each character, so a user can at least see the Unicode code point of the
character to be displayed.\footnote{\url{http://scripts.sil.org/UnicodeBMPFallbackFont}}

@@

A similar complaint to missing glyphs, discussed previously, is that while 
a glyph might be displayed, it does not look right. There are two
reasons for unexpected visual display, namely automatic font substitution and
faulty rendering. Like missing glyphs, any such problems are independent from
the Unicode Standard. The Unicode Standard only includes very general
information about characters and leaves the specific visual display to others to
decide on. Any faulty display is thus not to be blamed on the Unicode
Consortium, but on a complex interplay of different mechanisms happening in a
computer to turn Unicode code points into visual symbols. We will only sketch a
few aspects of this complex interplay here.

Most modern software applications (like Microsoft Word) offer some approach to
\textsc{automatic font substitution}. This means that when a text is written in
a specific font (e.g.~Times New Roman) and an inserted Unicode character does not
have a glyph within this font, then the software application will automatically
search for another font to display the glyph. The result will be that this
specific glyph will look slightly different from the others. This mechanism
works differently depending on the software application, only limited
user influence is usually expected and little feedback is given, which might be rather
frustrating to font-aware users.

@@

Another problem with visual display is related to so-called \textsc{font
rendering}. Font rendering refers to the process of the actual positioning of
Unicode characters on a page of written text. This positioning is actually a
highly complex challenge, and many things can go wrong in the process. Well-known
rendering difficulties, like proportional glyph size or ligatures, are reasonably
well understood by developers. Nevertheless	, the positioning of multiple diacritics relative to
a base character is still a widespread problem. Especially problematic is when 
more than one diacritic is supposed to be placed above (or
below) another. Even within the Latin script vertical placement 
often leads to unexpected effects in many modern software applications. 
The rendering problems arising in Arabic and in many scripts of Southeast
Asia (like Devanagari or Burmese) are even more complex.

To understand why any problems arise it is important to realize that there are
basically three different approaches to font rendering. The most widespread is
Adobe's and Microsoft's \textsc{OpenType} system. This approach makes it
relatively easy for font developers, but the font itself does not include all
details about the precise placement of individual characters. For those details,
additional script descriptions are necessary. All such systems can lead to
unexpected behavior.\footnote{For more details about OpenType, see
\url{http://www.adobe.com/products/type/opentype.html} and
\url{http://www.microsoft.com/typography/otspec/}. Additional systems for
complex text layout are, among others, Microsoft's DirectWrite
(\url{https://msdn.microsoft.com/library/dd368038.aspx}) and the open-source
project HarfBuzz (\url{http://www.freedesktop.org/wiki/Software/HarfBuzz/}).}
Alternative systems are \textsc{Apple Advanced Typography} (AAT) and the
open-source \textsc{Graphite} system produced and maintained by the Non-Roman Script Initiative of SIL International 
(SIL).\footnote{More information about AAT can be found at:
\url{https://developer.apple.com/fonts/}. \newline Graphite is described
in detail at:
\url{http://scripts.sil.org/default}.}
In these systems, a larger burden is placed on the description inside
the font.

@@

In other words, there are equivalent sequences of Unicode characters that should
be normalized, i.e.~transformed into a unique Unicode-sanctioned representation
of a character sequence called a \textsc{normalization form}. Unicode provides a
Unicode Normalization Algorithm, which essentially puts combining marks
into a specific logical order and it defines decomposition and composition
transformation rules to convert each string into one of four normalization
forms. We will discuss here the two most relevant normalization forms: NFC and
NFD.\@

>>>>

We also suggest installing at least one \textsc{fall-back
font}, which provides glyphs that show the user some information about
the underlying encoded character. Apple computers have such a font (which is
invisible to the user), which is designed by Michael Everson and made available
for other systems through the Unicode Consortium.\footnote{
\url{http://www.unicode.org/policies/lastresortfont\_eula.html}} Further, the
\textsc{GNU Unifont} is a clever way to produce bitmaps approximating the
intended glyph of each available character.\footnote{\url{http://unifoundry.com/unifont.html}} Finally,
SIL International provides a \textsc{SIL Unicode BMP Fallback
Font}. This font does not show a real 
glyph, but instead shows the hexadecimal code inside a box
for each character, so a user can at least see the Unicode code point of the
character intended for display.\footnote{\url{http://scripts.sil.org/UnicodeBMPFallbackFont}}

@@

A similar complaint to missing glyphs, discussed previously, is that while 
a glyph might be displayed, it does not look right. There are two
reasons for unexpected visual display, namely automatic font substitution and
faulty rendering. Like missing glyphs, any such problems are independent from
the Unicode Standard. The Unicode Standard only includes very general
information about characters and leaves the specific visual display for others to
decide on. Any faulty display is thus not to be blamed on the Unicode
Consortium, but on a complex interplay of different mechanisms happening in a
computer to turn Unicode code points into visual symbols. We will only sketch a
few aspects of this complex interplay here.

Most modern software applications (like Microsoft Word) offer some approach to
\textsc{automatic font substitution}. This means that when a text is written in
a specific font (e.g.~Times New Roman) and an inserted Unicode character does not
have a glyph within this font, then the software application will automatically
search for another font to display the glyph. The result will be that this
specific glyph will look slightly different from the others. This mechanism
works differently depending on the software application; only limited
user influence is usually expected and little feedback is given. This may be rather
frustrating to font-aware users.

@@

Another problem with visual display is related to so-called \textsc{font
rendering}. Font rendering refers to the process of the actual positioning of
Unicode characters on a page of written text. This positioning is actually a
highly complex challenge and many things can go wrong in the process. Well-known
rendering difficulties, like proportional glyph size or ligatures, are reasonably
well understood by developers. Nevertheless, the positioning of multiple diacritics relative to
a base character is still a widespread problem. Especially problematic is when 
more than one diacritic is supposed to be placed above (or
below) another. Even within the Latin script vertical placement 
often leads to unexpected effects in many modern software applications. 
The rendering problems arising in Arabic and in many scripts of Southeast
Asia (like Devanagari or Burmese) are even more complex.

To understand why these problems arise it is important to realize that there are
basically three different approaches to font rendering. The most widespread is
Adobe's and Microsoft's \textsc{OpenType} system. This approach makes it
relatively easy for font developers, but the font itself does not include all
details about the precise placement of individual characters. For those details,
additional script descriptions are necessary. All such systems can lead to
unexpected behavior.\footnote{For more details about OpenType, see
\url{http://www.adobe.com/products/type/opentype.html} and
\url{http://www.microsoft.com/typography/otspec/}. Additional systems for
complex text layout are, among others, Microsoft's DirectWrite
(\url{https://msdn.microsoft.com/library/dd368038.aspx}) and the open-source
project HarfBuzz (\url{http://www.freedesktop.org/wiki/Software/HarfBuzz/}).}
Alternative systems are \textsc{Apple Advanced Typography} (AAT) and the
open-source \textsc{Graphite} system produced and maintained by the Non-Roman Script Initiative of SIL International 
(SIL).\footnote{More information about AAT can be found at:
\url{https://developer.apple.com/fonts/}. \newline Graphite is described
in detail at:
\url{http://scripts.sil.org/default}.}
In these systems, a larger burden is placed on the description inside
the font.

@@

In other words, there are equivalent sequences of Unicode characters that should
be normalized, i.e.~transformed into a unique Unicode-sanctioned representation
of a character sequence called a \textsc{normalization form}. Unicode provides a
Unicode Normalization Algorithm, which puts combining marks
into a specific logical order and it defines decomposition and composition
transformation rules to convert each string into one of four normalization
forms. We will discuss here the two most relevant normalization forms: NFC and
NFD.

### 2018-06-23 book/chapters/preface.tex
ready for release, except doi

This text is meant as a practical guide for linguists and programmers, who work with data in multilingual computational environments. We introduce the basic concepts needed to understand how writing systems and character encodings function, and how they work together.

The intersection of the Unicode Standard and the International Phonetic Alphabet is often met with frustration by users. Nevertheless, the two standards have provided language researchers with a consistent computational architecture needed to process, publish and analyze data from many different languages. We bring to light common, but not always transparent, pitfalls that researchers face when working with Unicode and IPA.

We use quantitative methods in our research to compare languages to uncover and clarify their phylogenetic relationships. However, the majority of lexical data available from the world's languages is in author- or document-specific orthographies. Having identified and overcome the pitfalls involved in making writing systems and character encodings syntactically and semantically interoperable (to the extent that they can be), we created a suite of open-source Python and R software packages to work with languages using profiles that adequately describe their orthographic conventions. Using orthography profiles and these tools allows users to tokenize and transliterate text from diverse sources, so that they can be meaningfully compared and analyzed.

>>>>

This text is meant as a practical guide for linguists and programmers who work with data in multilingual computational environments. We introduce the basic concepts needed to understand how writing systems and character encodings function, and how they work together.

The intersection of the Unicode Standard and the International Phonetic Alphabet is often met with frustration by users. Nevertheless, the two standards have provided language researchers with the computational architecture needed to process, publish and analyze data from many different languages. We bring to light common, but not always transparent, pitfalls that researchers face when working with Unicode and IPA.

In our research, we use quantitative methods to compare languages to uncover and clarify their phylogenetic relationships. However, the majority of lexical data available from the world's languages is in author- or document-specific orthographies. Having identified and overcome the pitfalls involved in making writing systems and character encodings syntactically and semantically interoperable (to the extent that they can be), we have created a suite of open-source Python and R software packages to work with languages using profiles that adequately describe their orthographic conventions. Using these tools in combination with orthography profiles allows users to tokenize and transliterate text from diverse sources, so that they can be meaningfully compared and analyzed.

### 2018-06-23 book/chapters/unicode_approach.tex
ready for release, except doi

Each version of the Unicode Standard (\citealp{Unicode2018}, as of writing at
version 11) consists of a set of specifications and guidelines that include (i) a
core specification, (ii) code charts, (iii) standard annexes and (iv) a
character database.\footnote{All documents of the Unicode Standard are available
at: \url{http://www.unicode.org/versions/latest/}. For a quick survey of the use
of terminology inside the Unicode Standard, their glossary is particularly
useful, available at: \url{http://www.unicode.org/glossary/}. For a general
introduction to the principles of Unicode, Chapter 2 of the core specification,
called \textsc{general structure}, is particularly insightful. Unlike 
many other documents in the Unicode Standard, this general introduction is
relatively easy to read and illustrated with many interesting examples from
various orthographic traditions from all over the world.} The \textsc{core
specification} is a book aimed at human readers that describes the formal
standard for encoding multilingual text. The \textsc{code charts} provide a
human-readable online reference to the character contents of the Unicode
Standard in the form of PDF files. The \textsc{Unicode Standard Annexes (UAX)}
are a set of technical standards that describe the implementation of the Unicode
Standard for software development, web standards, and programming languages.
The \textsc{Unicode Character Database (UCD)} is a set of computer-readable text
files that describe the character properties, including a set of rich character
and writing system semantics, for each character in the Unicode Standard. In
this section, we introduce the basic Unicode concepts, but we will leave out
many details. Please consult the above-mentioned full documentation for a more
detailed discussion. Further note that the Unicode Standard is exactly that,
namely a standard. It normatively describes notions and rules to be followed. In
the actual practice of applying this standard in a computational setting, a
specific implementation is necessary. The most widely used implementation of the
Unicode Standard is the \textsc{International Components for Unicode (ICU)},
which offers C/C++ and Java libraries implementing the Unicode
Standard.\footnote{More information about the ICU is available here:
\url{http://icu-project.org}.}

@@

More difficult for text processing, because less standardized, is what the
Unicode Standard terms \textsc{tailored grapheme clusters}.\footnote{\url{http://unicode.org/reports/tr29/}} 
Tailored grapheme clusters are locale-dependent graphemes, i.e.~such combination of characters do
not function as text elements in all situations. Examples include the sequence
<c>~+~<h> for the Slovak digraph <ch> and the sequence <ky> in the Sisaala
practical orthography (pronounced as IPA /tʃ/,~\cite{Moran2006}). These grapheme
clusters are \textsc{tailored} in the sense that they must be specified on a
language-by-language or writing-system-by-writing-system basis. They are also 
grapheme clusters in these orthographies for processes such as collation (i.e.\ sorting).\footnote{\url{https://www.unicode.org/glossary/\#collation}}

>>>>

Each version of the Unicode Standard (\citealp{Unicode2018}, as of writing at
version 11.0.0) consists of a set of specifications and guidelines that include (i) a
core specification, (ii) code charts, (iii) standard annexes and (iv) a
character database.\footnote{All documents of the Unicode Standard are available
at: \url{http://www.unicode.org/versions/latest/}. For a quick survey of the use
of terminology inside the Unicode Standard, their glossary is particularly
useful, available at: \url{http://www.unicode.org/glossary/}. For a general
introduction to the principles of Unicode, Chapter 2 of the core specification,
called \textsc{general structure}, is particularly insightful. Unlike 
many other documents in the Unicode Standard, this general introduction is
relatively easy to read and illustrated with many interesting examples from
various orthographic traditions from all over the world.} The \textsc{core
specification} is a book aimed at human readers that describes the formal
standard for encoding multilingual text. The \textsc{code charts} provide a
human-readable online reference to the character contents of the Unicode
Standard in the form of PDF files. The \textsc{Unicode Standard Annexes (UAX)}
are a set of technical standards that describe the implementation of the Unicode
Standard for software development, web standards, and programming languages.
The \textsc{Unicode Character Database (UCD)} is a set of computer-readable text
files that describe the character properties, including a set of rich character
and writing system semantics, for each character in the Unicode Standard. In
this section, we introduce the basic Unicode concepts, but we will leave out
many details. Please consult the above-mentioned full documentation for a more
detailed discussion. Further note that the Unicode Standard is exactly that,
namely a standard. It normatively describes notions and rules to be followed. In
the actual practice of applying this standard in a computational setting, a
specific implementation is necessary. The most widely used implementation of the
Unicode Standard is the \textsc{International Components for Unicode (ICU)},
which offers C/C++ and Java libraries implementing the Unicode
Standard.\footnote{More information about the ICU is available here:
\url{http://icu-project.org}.}

@@

More difficult for text processing, because less standardized, is what the
Unicode Standard terms \textsc{tailored grapheme clusters}.\footnote{\url{http://unicode.org/reports/tr29/}} 
Tailored grapheme clusters are locale-dependent graphemes, i.e.~such combination of characters do
not function as text elements in all situations. Examples include the sequence
<c>~+~<h> for the Slovak digraph <ch> and the sequence <ky> in the Sisaala
practical orthography, which is pronounced as IPA /tʃ/ \citep{Moran2006}. These grapheme
clusters are \textsc{tailored} in the sense that they must be specified on a
language-by-language or writing-system-by-writing-system basis. They are also 
grapheme clusters in these orthographies for processes such as collation (i.e.\ sorting).\footnote{\url{https://www.unicode.org/glossary/\#collation}}
